### 2022-12-12 article/article.qmd
Update git repo

---
title: "Trade Dependence"
format: 
  html:
    self-contained: true
bibliography: bibliography.bib
execute: 
  message: false
  warning: false
  echo: false
---

```{r libraries}
library(tidyverse)
library(scales)
library(gtsummary)
library(modelsummary)
```

```{r data}
full_df <- rio::import(here::here("data", "full_df.csv"))
```

```{r models}
source(here::here("analysis", "models.R"))
```

Does trade dependence reduce conflict between states? Debate remains over the role and significance of economic interdependence in promoting peace: different scholars have found that dependence increases, decreases, and has no significant effect on the likelihood of conflict between states. However, this research has generally relied on a broad measure of bilateral trade dependence. This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence. If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption. Rarely, if ever, do we see states threaten to or impose total trade embargoes. Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy. Using a more disaggregated measure of trade dependence, I find that dependence does not have a significant effect on the likelihood of militarized disputes between states.

{{< pagebreak >}}

The commercial peace theory maintains that trade fosters peace between states because it makes conflict more costly. When a state is deciding whether to initiate conflict against another state, they weigh the benefits they expect to derive against the costs they expect to incur. When the benefits outweigh the costs, conflict begins. Commercial peace theory is a simple extension of this bargaining theory of conflict: the inevitable trade disruptions caused by conflict increase the costs states face, which decreases the likelihood that the benefits will outweigh the costs and conflict will begin, *ceteris paribus*.

Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation. Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error. Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP). However, a more disaggregated approach to trade dependence is worth exploring. We are interested in understanding how trade dependence can deter conflict. Rarely (if ever) do states threaten to sever their entire trade relationship with another state, including trade dependent states. Rather, we regularly see states threaten or enact trade bans on particular products. In fact, states often aim to threaten to, or actually, inflict the greatest harm on their rival whilst minimizing the costs they themselves will face. A broad understanding of trade dependence does not capture the real threat a potential aggressor faces or should expect to face. A more accurate measure of this factor is product-level trade dependence. Asymmetries at this level have been used as a foreign policy mechanism to deter or punish rival states.

Before we explore this dynamic in detail, let us step out in detail the process we seek better understand. First, tensions between two states build. This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an unfavourable bilateral foreign policy. The threatened state issues a formal request to the initiating state to change or reverse their policy actions. The initiating state refuses: the incentives for enacting the controversial policy continue to outweigh the cost of a soured relationship with the threatened state. The threatened state then states that it will adopt punitive policies towards the initiator, including trade embargoes on specific products on which the initiator depends. Here, the initiator's expected costs from pursuing this policy change: they know that if they continue along this path they will likely face economic costs in the form of trade disruptions. Does this deter them? Many scholars would ask: how costly is this disruption? This paper will answer this question. If the costs are not so great, the initiator may respond to this threat through further escalation, including through a militarized dispute. International relations scholarship uses this broad measure of conflict to include actions that sit below invasion or war. This is often the precursor to full-scale interstate war. States use these actions as an extreme signal of their intent to escalate a dispute to war, hoping to deter their opponent.

We are interested in the relationship between trade dependence and this pre-cursor to war. Certainly, if two states are at war with one another, we should expect trade to nearly, if not entirely, cease between them. I do not refute this. However, we should not expect trade to entirely cease between states when they are still circling each other, hoping to achieve their most desired policy outcome and avoid war. Tying a measure of the entire trade relationship to whether or not states initiate militarized interstate disputes (not war) against each other misses this important distinction. This paper seeks to address this.

The very broad measure of trade dependence dominant in the literature examines the total trade between states $i$ and $j$ relative to state $i$'s gross domestic product (GDP) in year $t$:

$$
TD_{ij,t} = \frac{M_{ij,t} + X_{ij,t}}{GDP_{i,t}}
$$

Some scholars argue that this measure captures the cost state $i$ considers when deciding whether to adopt some unfavorable foreign policy towards state $j$ [@oneal1997]. I disagree.

To understand the role that trade dependence plays in foreign policy formation, we not only need to further develop our understanding of trade dependence, but also of the realities faced by both the imposer and target of trade disruptions. Trade embargoes are costly for both the target and the imposer. Both of these actors will attempt to minimize the costs they will face and maximize those faced by their opponent. This occurs over a number of actions, with each actor gaining more information about their own and their opponent's tolerance to these economic costs as this bargaining game progresses.

Kavaklı et al [-@kavakli2020] provide a useful typology for understanding the type of actions each player can undertake. They argue that a state's reactions to trade disruptions can involve external and internal substitution. External substitution involves finding alternate trade partners for a product. Internal substitution involves locating new domestic supplies of the disrupted product. The less substitutions available to a state, the more vulnerable they will be to trade disruptions.

However, Gartzke and Westerwinter [-@gartzke2016] and Chen [-@chen2021] point out that we need to understand these asymmetries within the broader context of global trade. Within the network of product-level global trade, individual states will have symmetric and asymmetric relations with a number of different states.

## Trade dependence and militarized conflict

First, we need to understand how important trade between states $i$ and $j$ is to state $i$'s economy.

> H1: The greater the proportion of state \$i\$'s GDP derived from trade in product $p$ with state \$j\$, the less likely state $i$ is to initiate militarized conflict against state $j$.

Next, we need to understand whether state $i$ can locate alternative trade partners to state $j$ to mitigate the expected cost of trade disruptions from conflict.

> H2: The smaller the number of trade partners in the market for product $p$, the less likely state $i$ is to initiate militarized conflict against state $j$.

Next, we need to understand state $i$'s capacity for internal substitution. We will assume that state $i$ can divert exports in product $p$ to the domestic market to help mitigate costs from import disruptions imposed by state \$j\$. However, if state $i$ does not have a supply of product $p$ that it can divert from exports to fill the new gap in the domestic market, the greater the cost of trade disruptions in that product.

> H3: The smaller state $i$'s ratio of exports to imports in product $p$, the less likely state $i$ is to initiate militarized conflict against state $j$.

## Research design

### Scope

This analysis focuses on all directed state dyads in all years between 2000 and 2014. This includes `r nrow(scope) |> comma()` observations.

### Dependent variable

My dependent variable focuses on the onset of Militarized Interstate Disputes, or events in which one or more states threatens, displays, or uses force against another state or states. I include two binary dependent variables. First, I identify all dyad-years in which any MID was initiated. Second, I identify all dyad-years in which any MID involving at least one battle-related death was initiated. These variables are coded 1 if state $i$ initiated the MID against state $j$ in a given year. Otherwise, they receive a value of 0.

The data come from version five of Palmer et al's [-@palmer2022] Militarized Interstate Dispute data collection. This dataset includes all known MIDs from 1816-2014, recorded at the participant level. Between 2000 and 2014, `r mid_df |> filter(mid_initiated == 1) |> nrow() |> comma()` MIDs were initiated.

### Independent variables of interest

My primary independent variable of interest is trade dependence. This paper introduces a novel measure of trade dependence. Rather than looking at the balance of trade in total imports, exports, or trade value, I measure dependence in trade of specific products. This more fine-grained approach to measuring trade dependence allows us to examine more precisely the potential cost a foreign government could incur if their trade partner discovered their efforts at subversion.

This measure of trade dependence treats both the imposer and target of trade disrupts as complex actors. With regard to the imposer, it accounts for their attempts to disrupt trade in a product in which the trade asymmetries will result in the lowest cost to their economy.

With regard to the target, it captures the various responses available to them. It includes an understanding of how easily the target can locate an alternative trade partner for the disrupted product. If the target is able to easily and quickly locate an alternate trade partner, the threat or imposition of trade embargoes will have little deterrent effects.

This measure of trade dependence captures three components of a state's imports or exports: first, the importance of trade in the product to the state's economy; second, the diversity of the trade partner pool for each product; and third, each trade partner's dominance in the market for each product. I will expand on each of these components in turn.

First, how important is trade in each product to the state's economy? To build our understanding of a state's product-level reliance on trade with another state, we must first determine the degree to which their economy relies on access to or sales of that specific product. I, therefore, measure this as the value of trade in that product by the reporter as a proportion of its gross domestic product (GDP).

$$
r_{ij,p,t} = \frac{m_{ij,p,t} + x_{ij,p,t}}{GDP_{i,t}}
$$

Where $j$ is the trade partner and $p$ is the product. This value is measured for each year, $t$.

Second, we need to understand whether alternate trade partners are available for the state to turn to if trade in a product is disrupted. This is simply a measure of the number of other states trading in that product. I take the natural logarithum of this value to account for diminishing returns from many additional alternate trade partners.

Second, how diverse is the trade partner pool for each product? If the state only imports (or exports) a product from (to) one trade partner, they are more exposed than a state that trades in the product with many different trade partners. To measure this, I consider the value of trade with a partner as a proportion of the value of trade in the product.

For imports:

$$
d_{ij,p,t} = \frac{m_{ij,p,t}}{m_{ik,p,t}}
$$

For exports:

$$
d_{ij,p,t} = \frac{x_{ij,p,t}}{x_{ik,p,t}}
$$

Where $k$ represents all of state $i$'s trade partners.

Finally, how dominant is each trade partner in the market for each product? If the state trades in a product that is imported (or exported) by one trade partner, they are more vulnerable to trade disruptions from that one trade partner than if alternative trade partners exist.[^1] To measure this, I consider the value of products traded by the trade partner as a proportion of total value of that product traded.

[^1]: This, of course, assumes that the state can easily access these alternative markets. I leave it up to further research to capture the nuanced difficulties inherent in this process.

For imports:

$$
a_{j,p,t} = \frac{x_{jk,p,t}}{x_{p,t}}
$$

For exports:

$$
a_{j,p,t} = \frac{m_{jk,p,t}}{m_{p,t}}
$$

I then aggregate these measures to determine the level of trade dependence of state $i$ on their trade partner $j$ by product in a given year:

$$
dep_{ij,p,t} = \sum{r_{ij,p,t}, d_{ij,p,t}, a_{j,p,t}}
$$

This measure allows me to focus on trade in goods and services that are critical to the potential supporter. I argue that this represents the true cost that the potential sponsor considers when deciding whether or not to support a NSA group challenging a rival state.

### Previous measures of trade dependence

Chen [-@chen2021] develops a measure of Extended Dependence, or "the extent to which the potential challenger is economically dependent on the defensive allies of its potential target" (p.251). He defines this as:

$$
TD_{ij,t} = \frac{M_{ik,t} + X_{ik,t}}{GDP_{i,t}}
$$

Where $i$ is the potential challenger, $j$ is the target, $k$ is the target's allies, and $t$ is the given year. He lags this value by one year to address concerns surrounding endogeneity and reverse causality.

### Control variables

### 2022-12-12 analysis/models.R

@@ -1,0 +1,47 @@
+library(tidyverse)
+library(broom)
+library(gtsummary)
+
+# Load in full dataset ----------------------------------------------------
+
+full_df <- rio::import(here::here("data", "full_df.csv"))
+
+# Basic models ------------------------------------------------------------
+
+m1 <- glm(
+  mid_onset_all ~ trade_dep_total, 
+  family = binomial(link = "logit"), 
+  data = full_df
+)
+
+t1 <- tbl_regression(m1, exponentiate = T)
+
+m2 <- glm(
+  mid_onset_all ~ trade_dep_broad, 
+  family = binomial(link = "logit"), 
+  data = full_df
+)
+
+t2 <- tbl_regression(m2, exponentiate = T)
+
+tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
+
+# Controlled models -------------------------------------------------------
+
+m3 <- glm(
+  mid_onset_all ~ trade_dep_total + conttype + v2x_polyarchy1 + v2x_polyarchy2 + v2x_polyarchy, 
+  family = binomial(link = "logit"), 
+  data = full_df
+)
+
+t3 <- tbl_regression(m3, exponentiate = T)
+
+m4 <- glm(
+  mid_onset_all ~ trade_dep_broad + conttype + v2x_polyarchy1 + v2x_polyarchy2 + v2x_polyarchy, 
+  family = binomial(link = "logit"), 
+  data = full_df
+)
+
+t4 <- tbl_regression(m4, exponentiate = T)
+
+tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))

### 2022-12-12 data-processing/data-processing.R

@@ -1,0 +1,100 @@
+library(tidyverse)
+library(wbstats)
+library(countrycode)
+library(peacesciencer)
+
+# Scope -------------------------------------------------------------------
+
+scope <- create_dyadyears(system = "cow", mry = F, directed = T) |> 
+  filter(year == 2014, !ccode1 %in% c(347, 713)) |> 
+  mutate(
+    country1 = countrycode(ccode1, "cown", "country.name"),
+    country2 = countrycode(ccode2, "cown", "country.name")
+  )
+
+# MID ---------------------------------------------------------------------
+# Note, both sides can initiiate the war. 
+mid_df <- scope |> 
+  add_cow_mids(keep = c("dispnum", "sidea1", "sidea2", "fatality", "orig1", "orig2")) |> 
+  transmute(
+    ccode1,
+    country1,
+    ccode2,
+    country2,
+    year,
+    mid_onset_all = if_else(cowmidonset == 1 & orig1 == 1, 1, 0)
+  ) |> 
+  mutate(mid_onset_all = factor(mid_onset_all))
+
+# Trade dependence --------------------------------------------------------
+
+trade_raw <- rio::import(here::here("data-raw", "trade_data_2014.csv")) |> 
+  janitor::clean_names() |> 
+  transmute(
+    year = yr, 
+    country1 = countrycode(rt_title, "country.name", "country.name"), 
+    country2 = countrycode(pt_title, "country.name", "country.name",
+                           custom_match = c("World" = "World")), 
+    rg_desc, 
+    cmd_code, 
+    cmd_desc_e, 
+    trade_value
+  ) |> 
+  filter(rg_desc %in% c("Export", "Import")) |> 
+  as_tibble()
+
+trade_bilat <- trade_raw |> 
+  filter(country2 != "World")
+
+gdp_df <- wbstats::wb_data("NY.GDP.MKTP.CD", start_date = 2014, end_date = 2014, return_wide = F) |> 
+  transmute(
+    country1 = countrycode(iso3c, "iso3c", "country.name"), 
+    year = date, 
+    gdp = value
+  )
+
+trade_dep <- trade_bilat |> 
+  group_by(year, country1, country2, cmd_desc_e) |> 
+  summarise(trade_value_total = sum(trade_value, na.rm = T)) |> 
+  left_join(gdp_df, by = c("year", "country1")) |> 
+  mutate(trade_dep_total = (trade_value_total / gdp) * 100) |> 
[…]

### 2022-12-12 src/src_trade.R

@@ -1,0 +1,83 @@
+library(tidyverse)
+library(countrycode)
+library(httr)
+library(jsonlite)
+
+reporter_ids <- fromJSON("https://comtrade.un.org/Data/cache/reporterAreas.json")$results |> 
+  mutate(ccode = countrycode(text, "country.name", "cown")) |> 
+  drop_na(ccode) |> 
+  filter(!text %in% c("East and West Pakistan",
+                      "Fmr Dem. Rep. of Germany",
+                      "Fmr Fed. Rep. of Germany",
+                      "India, excl. Sikkim",
+                      "Peninsula Malaysia",
+                      "Fmr Panama-Canal-Zone",
+                      "Fmr Panama, excl.Canal Zone",
+                      "Fmr USSR",
+                      "Saint Kitts, Nevis and Anguilla",
+                      "Fmr Sudan",
+                      "US Virgin Isds",
+                      "USA (before 1981)",
+                      "Fmr Dem. Rep. of Vietnam",
+                      "Fmr Rep. of Vietnam",
+                      "Fmr Yugoslavia",
+                      "Fmr Rhodesia Nyas",
+                      "Fmr Ethiopia")) |> 
+  pull(id)
+
+pull_scope <- tibble(
+  year = 2008:2013
+) |> 
+  full_join(tibble(reporter_id = reporter_ids), by = character()) |> 
+  arrange(desc(year))
+
+pull_comtrade <- function(reporter_id, year) {
+  
+  query <- glue::glue("https://comtrade.un.org/api/get?r={ reporter_id }&ps={ year }&fmt=json&freq=A&head=M&px=HS&cc=AG2")
+  
+  result <- tryCatch(
+    fromJSON(query),
+    error = function(err) {
+      
+      print(glue::glue("Limit reached, wait one hour from { lubridate::now() }"))
+      Sys.sleep(3600)
+      return(fromJSON(query))
+      
+    }
+  )
+  
+  print(glue::glue("Completing { reporter_id } in { year }"))
+  print(result$validation$status$name)
+  
+  if (result$validation$status$name != "Result too large" & length(result$dataset) > 0) {
+    
+    return(
+      tibble(result$dataset)
+    )
+    
+  } else if (result$validation$status$name != "Result too large" & length(result$dataset) == 0) {
+    
+    return(
+      tibble(
+        rtCode = reporter_id,
+        status = "No results found"
+      )
+    )
[…]

### 2022-12-12 src/src_trade_large_countries.R

@@ -1,0 +1,159 @@
+library(tidyverse)
+library(countrycode)
+library(httr)
+library(jsonlite)
+
+# Read in existing trade data
+annual_trade_df <- rio::import(here::here("data-raw", "trade_data_2014.csv"))
+
+# Attempt direction-based pull --------------------------------------------
+
+# Find countries for which yearly trade data is too big
+pull_scope <- tibble(direction = c(1, 2)) |> 
+  full_join(
+    rio::import(here::here("data-raw", "trade_data_2014.csv")) |> 
+      filter(status == "Result too large") |> 
+      select(rtCode),
+    by = character()
+  )
+
+pull_large_comtrade <- function(reporter_id, direction) {
+  
+  query <- glue::glue("https://comtrade.un.org/api/get?r={ reporter_id }&ps=2014&rg={ direction }&fmt=json&freq=A&head=M&px=HS&cc=AG2")
+  
+  result <- tryCatch(
+    fromJSON(query),
+    error = function(err) {
+      
+      print(glue::glue("Limit reached, wait one hour from { lubridate::now() }"))
+      Sys.sleep(3600)
+      return(fromJSON(query))
+      
+    }
+  )
+  
+  print(glue::glue("Completing { reporter_id } for { direction }"))
+  print(result$validation$status$name)
+  
+  if (result$validation$status$name != "Result too large" & length(result$dataset) > 0) {
+    
+    return(
+      tibble(result$dataset)
+    )
+    
+  } else if (result$validation$status$name != "Result too large" & length(result$dataset) == 0) {
+    
+    return(
+      tibble(
+        rtCode = reporter_id,
+        status = glue::glue("No results found: { direction }")
+      )
+    )
+    
+  } else {
+    
+    return(
+      tibble(
+        rtCode = as.integer(reporter_id),
+        rgCode = as.integer(direction),
+        status = glue::glue("Results too large: { direction }")
+      )
+    )
+    
+  }
+  
+}
+
+safely_pull_large_comtrade <- safely(~ pull_large_comtrade(.x, .y))
+
+large_trade_df <- map2_dfr(
+  pull_scope$rtCode, 
+  pull_scope$direction, 
+  ~ safely_pull_large_comtrade(.x, .y)
+)
+
+# Update results
+large_annual_trade_df <- annual_trade_df |> 
+  as_tibble() |> 
[…]

### 2022-12-12 article/article.qmd
Remove references to more complex trade dep measure

@@
 
-To understand the role that trade dependence plays in foreign policy formation, we not only need to further develop our understanding of trade dependence, but also of the realities faced by both the imposer and target of trade disruptions.
-Trade embargoes are costly for both the target and the imposer.
-Both of these actors will attempt to minimize the costs they will face and maximize those faced by their opponent.
-This occurs over a number of actions, with each actor gaining more information about their own and their opponent's tolerance to these economic costs as this bargaining game progresses.
-
-Kavaklı et al [-@kavakli2020] provide a useful typology for understanding the type of actions each player can undertake.
-They argue that a state's reactions to trade disruptions can involve external and internal substitution.
-External substitution involves finding alternate trade partners for a product.
-Internal substitution involves locating new domestic supplies of the disrupted product.
-The less substitutions available to a state, the more vulnerable they will be to trade disruptions.
-
-However, Gartzke and Westerwinter [-@gartzke2016] and Chen [-@chen2021] point out that we need to understand these asymmetries within the broader context of global trade.
-Within the network of product-level global trade, individual states will have symmetric and asymmetric relations with a number of different states.
-
 ## Trade dependence and militarized conflict
-
-First, we need to understand how important trade between states $i$ and $j$ is to state $i$'s economy.
-
-> H1: The greater the proportion of state \$i\$'s GDP derived from trade in product $p$ with state \$j\$, the less likely state $i$ is to initiate militarized conflict against state $j$.
-
-Next, we need to understand whether state $i$ can locate alternative trade partners to state $j$ to mitigate the expected cost of trade disruptions from conflict.
-
-> H2: The smaller the number of trade partners in the market for product $p$, the less likely state $i$ is to initiate militarized conflict against state $j$.
-
-Next, we need to understand state $i$'s capacity for internal substitution.
-We will assume that state $i$ can divert exports in product $p$ to the domestic market to help mitigate costs from import disruptions imposed by state \$j\$.
-However, if state $i$ does not have a supply of product $p$ that it can divert from exports to fill the new gap in the domestic market, the greater the cost of trade disruptions in that product.
-
-> H3: The smaller state $i$'s ratio of exports to imports in product $p$, the less likely state $i$ is to initiate militarized conflict against state $j$.
 

### 2022-12-12 article/article.qmd
Remove references to more complex trade dep measure

--- 2022-12-12 article/article.qmd
@@
 This occurs over a number of actions, with each actor gaining more information about their own and their opponent's tolerance to these economic costs as this bargaining game progresses.
-
-Kavaklı et al [-@kavakli2020] provide a useful typology for understanding the type of actions each player can undertake.
-They argue that a state's reactions to trade disruptions can involve external and internal substitution.
-External substitution involves finding alternate trade partners for a product.
-Internal substitution involves locating new domestic supplies of the disrupted product.
-The less substitutions available to a state, the more vulnerable they will be to trade disruptions.
-
-However, Gartzke and Westerwinter [-@gartzke2016] and Chen [-@chen2021] point out that we need to understand these asymmetries within the broader context of global trade.
-Within the network of product-level global trade, individual states will have symmetric and asymmetric relations with a number of different states.
 

### 2022-12-12 article/article.qmd
Update scope description

--- 2022-12-12 article/article.qmd
@@
 
-The very broad measure of trade dependence dominant in the literature examines the total trade between states $i$ and $j$ relative to state $i$'s gross domestic product (GDP) in year $t$:
+## Trade dependence and militarized conflict
 
-$$
-TD_{ij,t} = \frac{M_{ij,t} + X_{ij,t}}{GDP_{i,t}}
-$$
+Returning to our motivating question: does trade dependence deter states from initiating militarized conflict against each other?
+The commercial peace theory argues that trade dependence increases the expected cost of initiating conflict against a trade partner, decreasing the expected net benefits derived from such an action and, therefore, reducing the likelihood a state will initiate conflict.
+This leads to the following testable hypothesis:
 
-Some scholars argue that this measure captures the cost state $i$ considers when deciding whether to adopt some unfavorable foreign policy towards state $j$ [@oneal1997]. I disagree.
-
-## Trade dependence and militarized conflict
+> H1: As product-level trade dependence between two states increases, the likelihood that a state will initiate militarized conflict against its trade partner decreases.
 
@@
 
-This analysis focuses on all directed state dyads in all years between 2000 and 2014. This includes `r nrow(scope) |> comma()` observations.
+This analysis focuses on all directed state dyads in 2014.
+This includes `r nrow(scope) |> comma()` observations.
+We are interested in whether or not a trade dependent state initiates conflict against another state.
+This is a directed question.
+We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on their other.
+
+The narrow timeframe included in this analysis is driven by data constraints.
+I only have access to complete trade and militarized interstate dispute data for 2014.
+A natural first step for extending this analysis is including a broader timeframe in the analysis.
 
@@
 
-My dependent variable focuses on the onset of Militarized Interstate Disputes, or events in which one or more states threatens, displays, or uses force against another state or states.
+My dependent variable focuses on the onset of militarized interstate disputes, or events in which one or more states threaten, display, or use force against another state or states.
 I include two binary dependent variables.

### 2022-12-12 article/article.qmd
Update MID data description to include all MIDs not fatal MIDs

@@
 
-My dependent variable focuses on the onset of militarized interstate disputes, or events in which one or more states threaten, display, or use force against another state or states.
-I include two binary dependent variables.
-First, I identify all dyad-years in which any MID was initiated.
-Second, I identify all dyad-years in which any MID involving at least one battle-related death was initiated.
-These variables are coded 1 if state $i$ initiated the MID against state $j$ in a given year.
-Otherwise, they receive a value of 0.
+My dependent variable focuses on the onset of militarized interstate disputes (MIDs), or events in which one or more states threaten, display, or use force against another state or states.
+If state $i$ initiated the MID against state $j$ in a given year, this variable is coded 1.
+Otherwise, it receives a value of 0.
 
@@
 This dataset includes all known MIDs from 1816-2014, recorded at the participant level.
-Between 2000 and 2014, `r mid_df |> filter(mid_initiated == 1) |> nrow() |> comma()` MIDs were initiated.
+In 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.
 

### 2022-12-12 data-processing/data-processing.R
Update to include cmd descriptions

@@ -2,6 +2,7 @@ library(tidyverse)
 library(wbstats)
 library(countrycode)
 library(peacesciencer)
+library(jsonlite)
 
 # Scope -------------------------------------------------------------------
 
@@ -53,15 +54,23 @@ gdp_df <- wbstats::wb_data("NY.GDP.MKTP.CD", start_date = 2014, end_date = 2014,
     gdp = value
   )
 
+commodities <- fromJSON("https://comtrade.un.org/Data/cache/classificationHS.json")$results |> 
+  filter(parent == "TOTAL") |> 
+  transmute(
+    cmd_code = as.numeric(id),
+    cmd_desc_e = str_sub(text, 6)
+  )
+
 trade_dep <- trade_bilat |> 
-  group_by(year, country1, country2, cmd_desc_e) |> 
+  group_by(year, country1, country2, cmd_code) |> 
   summarise(trade_value_total = sum(trade_value, na.rm = T)) |> 
+  left_join(commodities, by = "cmd_code") |> 
   left_join(gdp_df, by = c("year", "country1")) |> 
-  mutate(trade_dep_total = (trade_value_total / gdp) * 100) |> 
+  mutate(trade_dep_total = (trade_value_total / gdp)) |> 
   group_by(country1, country2, year) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
-  select(year:cmd_desc_e, trade_dep_total)
+  select(year:cmd_code, cmd_desc_e, trade_dep_total)
 
 trade_dep_broad <- trade_bilat |> 
   group_by(year, country1, country2) |> 
@@ -71,7 +80,7 @@ trade_dep_broad <- trade_bilat |>
     country1, 
     country2,
     year,
-    trade_dep_broad = (total_trade_value / gdp) * 100
+    trade_dep_broad = (total_trade_value / gdp)
   ) |> 
   ungroup()
 
@@ -93,7 +102,8 @@ controls <- scope |>
 full_df <- mid_df |> 
   left_join(trade_dep, by = c("country1", "country2", "year")) |> 
   left_join(trade_dep_broad, by = c("country1", "country2", "year")) |>
-  left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year"))
+  left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year")) |> 
[…]

### 2022-12-12 article/article.qmd
Write up results for max trade dep

@@
 source(here::here("analysis", "models.R"))
+```
+
+```{r helpers}
+theme_custom <- function() {
+
+  theme_minimal() + 
+  theme(plot.title.position = "plot",
+        panel.grid = element_blank(),
+        plot.title = element_text(face = "bold"),
+        legend.position = "bottom")
+
+}
 ```
@@
 Certainly, if two states are at war with one another, we should expect trade to nearly, if not entirely, cease between them.
-I do not refute this.
 However, we should not expect trade to entirely cease between states when they are still circling each other, hoping to achieve their most desired policy outcome and avoid war.
@@
 
-My primary independent variable of interest is trade dependence.
-This paper introduces a novel measure of trade dependence.
+My primary independent variable of interest is bilateral trade dependence.
+This paper introduces a novel measure of this dependence.
 Rather than looking at the balance of trade in total imports, exports, or trade value, I measure dependence in trade of specific products.
-This more fine-grained approach to measuring trade dependence allows us to examine more precisely the potential cost a foreign government could incur if their trade partner discovered their efforts at subversion.
+This more fine-grained approach to measuring trade dependence allows us to examine more precisely the potential costs an initiator expects to incur if they escalate a dispute to a MID.
 
-This measure of trade dependence treats both the imposer and target of trade disrupts as complex actors.
-With regard to the imposer, it accounts for their attempts to disrupt trade in a product in which the trade asymmetries will result in the lowest cost to their economy.
+To do this, I adopt the classical structure proposed by Oneal and Russett [-@oneal1997] but adapt it to trade in each product.
+I use bilateral annual export and import data provided by UN Comtrade.
+Commodities are grouped according to the standard two digit harmonized system.
+For example, all live animals, including horses, bovines, swine, sheep, and goats, are grouped together into a category called *animals - live*.
+My theory assumes a degree of substitution within this group.
+It would be interesting to explore the level of granularity required to meaningfully make this assumption.
+Sadly, this is difficult to test because we lose data completeness as we increase granularity.
 
-With regard to the target, it captures the various responses available to them.
-It includes an understanding of how easily the target can locate an alternative trade partner for the disrupted product.
-If the target is able to easily and quickly locate an alternate trade partner, the threat or imposition of trade embargoes will have little deterrent effects.
-
-This measure of trade dependence captures three components of a state's imports or exports: first, the importance of trade in the product to the state's economy; second, the diversity of the trade partner pool for each product; and third, each trade partner's dominance in the market for each product.
-I will expand on each of these components in turn.
-
-First, how important is trade in each product to the state's economy?
-To build our understanding of a state's product-level reliance on trade with another state, we must first determine the degree to which their economy relies on access to or sales of that specific product.
-I, therefore, measure this as the value of trade in that product by the reporter as a proportion of its gross domestic product (GDP).
+Formally, my measure of trade dependence is defined as the total imports and exports from state $i$ to (or from) state $j$ for a given commodity in a given year relative to state $i$'s GDP for that year.
 
@@
 
-Second, we need to understand whether alternate trade partners are available for the state to turn to if trade in a product is disrupted.
-This is simply a measure of the number of other states trading in that product.
-I take the natural logarithum of this value to account for diminishing returns from many additional alternate trade partners.
+```{r}
+trade_dep_max <- full_df |> 
+  slice_max(trade_dep_total, with_ties = F) |> 
+  mutate(cmd_desc_e = str_to_lower(cmd_desc_e))
+```
 
-Second, how diverse is the trade partner pool for each product?
-If the state only imports (or exports) a product from (to) one trade partner, they are more exposed than a state that trades in the product with many different trade partners.
-To measure this, I consider the value of trade with a partner as a proportion of the value of trade in the product.
+In 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
+The most dependent pairing is between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`: `r pull(trade_dep_max, country1)`'s trade in `r pull(trade_dep_max, cmd_desc_e)` with `r pull(trade_dep_max, country2)` amounts to `r trade_dep_max |> pull(trade_dep_total) |> percent()` of `r pull(trade_dep_max, country1)`'s GDP in 2014.
 
-For imports:
+States trade in a large number of commodities.
+For each directed state dyad, we get the same number of measures of dependence as products traded between them.
+We need to reduce this to a single measure of dependence.
+I propose five options: the minimum, the 25th percentile, the average, the 75th percentile, and the maximum level of trade dependence between two states.
+Let's discuss the logic of the extremes - the minimum and the maximum values first.
+The minimum value represents the lowest possible cost state $j$ could threaten to impose on state $i$ in a given year.
+If this is relatively high, state $i$ is probably very trade dependent on state $j$.
+This should be a very strong measure of dependence.
+However, state $j$ is unlikely to threaten to stop trade in this commodity.
+Remember, they aim to impose the greatest possible cost on state $i$ whilst minimizing the cost to themselves.
+Perhaps a better candidate is trade in the most dependent commodity.
+To qualify this measure, I will also test intermediate measures of dependence: the 25th, 50th, and 75th percentiles of trade dependence across all traded commodities.
+These measures perhaps better capture the dynamics at play with the extreme measures whilst allowing state $j$ to minimize the cost it bears and also to select a number of commodities to disrupt, not just one.
+To illustrate, let's continue to examine the trade relationship between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`.
 
-$$
-d_{ij,p,t} = \frac{m_{ij,p,t}}{m_{ik,p,t}}
-$$
+```{r}
+trade_dep |> 
+  filter(
+      country1 == trade_dep_max$country1,
+      country2 == trade_dep_max$country2,
+      trade_dep_total > 0
+  ) |> 
+  summarise(
+    `Trade dependence` = round(quantile(trade_dep_total) * 100, 3)
+  ) |> 
+  mutate(
+    Level = c("Minimum", "25th percentile", "Mean", "75th percentile", "Maximum")
+  ) |> 
+  relocate(Level) |> 
+  knitr::kable()
+```
 
-For exports:
-
-$$
-d_{ij,p,t} = \frac{x_{ij,p,t}}{x_{ik,p,t}}
-$$
-
-Where $k$ represents all of state $i$'s trade partners.
-
-Finally, how dominant is each trade partner in the market for each product?
-If the state trades in a product that is imported (or exported) by one trade partner, they are more vulnerable to trade disruptions from that one trade partner than if alternative trade partners exist.[^1] To measure this, I consider the value of products traded by the trade partner as a proportion of total value of that product traded.
-
-[^1]: This, of course, assumes that the state can easily access these alternative markets. I leave it up to further research to capture the nuanced difficulties inherent in this process.
-
-For imports:
-
-$$
-a_{j,p,t} = \frac{x_{jk,p,t}}{x_{p,t}}
-$$
-
-For exports:
-
-$$
-a_{j,p,t} = \frac{m_{jk,p,t}}{m_{p,t}}
-$$
-
-I then aggregate these measures to determine the level of trade dependence of state $i$ on their trade partner $j$ by product in a given year:
-
-$$
-dep_{ij,p,t} = \sum{r_{ij,p,t}, d_{ij,p,t}, a_{j,p,t}}
-$$
-
-This measure allows me to focus on trade in goods and services that are critical to the potential supporter.
-I argue that this represents the true cost that the potential sponsor considers when deciding whether or not to support a NSA group challenging a rival state.
-
-### Previous measures of trade dependence
-
-Chen [-@chen2021] develops a measure of Extended Dependence, or "the extent to which the potential challenger is economically dependent on the defensive allies of its potential target" (p.251).
-He defines this as:
-
-$$
-TD_{ij,t} = \frac{M_{ik,t} + X_{ik,t}}{GDP_{i,t}}
-$$
-
-Where $i$ is the potential challenger, $j$ is the target, $k$ is the target's allies, and $t$ is the given year.
-He lags this value by one year to address concerns surrounding endogeneity and reverse causality.
+This measure allows me to focus on trade in goods that are critical to the potential initiator.
+I argue that this better represents the true cost they expect to incur if they escalate tensions with a trade partner to a MID.
 
 ### Control variables
+
+Following other literature to enhance comparability, I include the following controls.
+
+## Results
+
+Let's look at the relationship between the predicted probability that state $i$ will initiate a MID with state $j$ for a given level of product-level trade dependence.
+First, I will explore the different degrees of effects predicted using the different measures of product-level trade dependence.
+I will then compare the relationship between MID initiation and product-level and broad, total trade dependence.
+Finally, I will include traditional controls for a state's decision to initiate a MID against a rival to see whether trade dependence is a significant factor in this complex decision.
+
+We will compare this to the predicted probability derived using the traditional broad measure of trade dependence.
+
+```{r}
+tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
+```
+
+I used a binary logistic regression model to predict the probability that state $i$ initiates a MID against state $j$ for a given level of trade dependence.
+Using product-level trade dependence, I find that increasing dependence by one percentage point **increases** the odds that state $i$ will initiate MID against state $j$ by `r tidy(m1, exponentiate = T) |> filter(term == "trade_dep_total") |> pull(estimate) |> round(2)` on average.
+This relationship is highly statistically significant.
+This more fine-grained measure of trade dependence predicts a stronger relationship between conflict and trade dependence than the more tradition and broad-based approach.
+Using total trade dependence, I find that increasing dependence by one percentage point increases the odds that state $i$ will initiate MID against state $j$ by `r tidy(m2, exponentiate = T) |> filter(term == "trade_dep_broad") |> pull(estimate) |> round(2)` on average.
+
+*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^3]
+
+[^3]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.
+
+```{r}
+ggplot(pred_m1_m2, aes(x = trade_dep, y = .fitted, colour = type)) + 
+  geom_line() + 
+  theme_custom() + 
+  labs(
+    title = "Predicted probability of MID given product-level trade dependence",
+    x = "Trade dependence (%)",
+    y = NULL,
+    colour = "Measure"
+  )
+```
+
+It is worth noting that product-level trade dependence predicts a much more aggressive relationship between conflict and trade dependence: the point at which MIDs are predicted to be more likely than not is at `r pred_m1_m2 |> filter(type == "Product-level", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for product-level trade dependence and `r pred_m1_m2 |> filter(type == "Total trade", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for total trade dependence.
+
+Moving from these base models to ones that incorporate traditional controls for MIDs, we learn that the relationship between conflict and trade dependence is not significant.
+
+```{r}
+tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
+```
+
+For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds.
+Significant determinants of conflict include contiguity and state $i$'s level of democracy.
+Concentrating on product-level trade dependence, I find that state $i$ is less to initiate an MID against a state with which it shares a land border.
+I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$.
+Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID, holding all else constant.

### 2022-12-12 analysis/models.R

@@ -1,47 +1,63 @@
 library(tidyverse)
 library(broom)
 library(gtsummary)
+library(labelled)
 
 # Load in full dataset ----------------------------------------------------
 
-full_df <- rio::import(here::here("data", "full_df.csv"))
+perc_df <- rio::import(here::here("data", "full_df.csv")) |> 
+  mutate(across(trade_dep_total:trade_dep_broad, ~ .x * 100)) |> 
+  set_variable_labels(trade_dep_total = "Trade dependence, product-level",
+                      trade_dep_broad = "Trade dependence, total trade", 
+                      conttype = "Contiguous",
+                      v2x_polyarchy1 = "Democracy, state i",
+                      v2x_polyarchy2 = "Democracy, state j")
 
 # Basic models ------------------------------------------------------------
 
 m1 <- glm(
   mid_onset_all ~ trade_dep_total, 
   family = binomial(link = "logit"), 
-  data = full_df
+  data = perc_df
 )
 
-t1 <- tbl_regression(m1, exponentiate = T)
+t1 <- tbl_regression(m1, exponentiate = T, intercept = T)
 
 m2 <- glm(
   mid_onset_all ~ trade_dep_broad, 
   family = binomial(link = "logit"), 
-  data = full_df
+  data = perc_df
 )
 
-t2 <- tbl_regression(m2, exponentiate = T)
+t2 <- tbl_regression(m2, exponentiate = T, intercept = T)
 
-tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
+pred_m1_m2 <- augment(
+  m1, newdata = tibble(trade_dep_total = seq(0, 100, 1)), type.predict = "response"
+) |> 
+  mutate(type = "Product-level") |> 
+  rename(trade_dep = trade_dep_total) |> 
+  bind_rows(
+    augment(
+      m2, newdata = tibble(trade_dep_broad = seq(0, 100, 1)), type.predict = "response"
+    ) |>
+      mutate(type = "Total trade") |> 
+      rename(trade_dep = trade_dep_broad)
+  )
 
 # Controlled models -------------------------------------------------------
 
 m3 <- glm(
-  mid_onset_all ~ trade_dep_total + conttype + v2x_polyarchy1 + v2x_polyarchy2 + v2x_polyarchy, 
[…]

### 2022-12-12 data-processing/data-processing.R

@@ -59,7 +59,8 @@ commodities <- fromJSON("https://comtrade.un.org/Data/cache/classificationHS.jso
   transmute(
     cmd_code = as.numeric(id),
     cmd_desc_e = str_sub(text, 6)
-  )
+  ) |> 
+  distinct()
 
 trade_dep <- trade_bilat |> 
   group_by(year, country1, country2, cmd_code) |> 
@@ -67,10 +68,13 @@ trade_dep <- trade_bilat |>
   left_join(commodities, by = "cmd_code") |> 
   left_join(gdp_df, by = c("year", "country1")) |> 
   mutate(trade_dep_total = (trade_value_total / gdp)) |> 
+  ungroup()
+
+trade_dep_max <- trade_dep |> 
   group_by(country1, country2, year) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
-  select(year:cmd_code, cmd_desc_e, trade_dep_total)
+  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
 
 trade_dep_broad <- trade_bilat |> 
   group_by(year, country1, country2) |> 
@@ -100,7 +104,7 @@ controls <- scope |>
 # Create full dataset -----------------------------------------------------
 # TODO: Work out what are true NAs and what are 0s. 
 full_df <- mid_df |> 
-  left_join(trade_dep, by = c("country1", "country2", "year")) |> 
+  left_join(trade_dep_max, by = c("country1", "country2", "year")) |> 
   left_join(trade_dep_broad, by = c("country1", "country2", "year")) |>
   left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year")) |> 
   mutate(across(trade_dep_total:trade_dep_broad, ~ replace_na(.x, 0)))

### 2022-12-12 article/article.qmd
Update footnote on different aggregations of trade dep

@@
 This analysis focuses on all directed state dyads in 2014.
-This includes `r nrow(scope) |> comma()` observations.
+This includes `r nrow(full_df) |> comma()` observations.
 We are interested in whether or not a trade dependent state initiates conflict against another state.
@@
 We need to reduce this to a single measure of dependence.
-I propose five options: the minimum, the 25th percentile, the average, the 75th percentile, and the maximum level of trade dependence between two states.
-Let's discuss the logic of the extremes - the minimum and the maximum values first.
+I use the product with the highest value of dependence between states $i$ and $j$.
+State $j$ will strategically select the product or products that maximize the cost state $i$ should expect to incur in retaliation for escalating tensions.
+The trade in a product on which state $i$ is most dependent is a good candidate for this strategic threat or imposition of a trade embargo.[^1]
+
+[^1]: I tested five options for this aggregation: the minimum, the 25th percentile, the average, the 75th percentile, and the maximum level of trade dependence between two states.
+Let's first discuss the logic of the extremes: the minimum and maximum values.
 The minimum value represents the lowest possible cost state $j$ could threaten to impose on state $i$ in a given year.
@@
 Perhaps a better candidate is trade in the most dependent commodity.
-To qualify this measure, I will also test intermediate measures of dependence: the 25th, 50th, and 75th percentiles of trade dependence across all traded commodities.
+To qualify these measures, I also also tested intermediate measures of dependence: the 25th, 50th, and 75th percentiles of trade dependence across all traded commodities.
 These measures perhaps better capture the dynamics at play with the extreme measures whilst allowing state $j$ to minimize the cost it bears and also to select a number of commodities to disrupt, not just one.
-To illustrate, let's continue to examine the trade relationship between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`.
-
-```{r}
-trade_dep |> 
-  filter(
-      country1 == trade_dep_max$country1,
-      country2 == trade_dep_max$country2,
-      trade_dep_total > 0
-  ) |> 
-  summarise(
-    `Trade dependence` = round(quantile(trade_dep_total) * 100, 3)
-  ) |> 
-  mutate(
-    Level = c("Minimum", "25th percentile", "Mean", "75th percentile", "Maximum")
-  ) |> 
-  relocate(Level) |> 
-  knitr::kable()
-```
+I found that the maximum, 75th percentile, and average measures of trade dependence were statistically significant to initiation of MIDs.
+In fact, the average measure of trade dependence produced the strongest effect on MID initiation.
+However, I am not sure that these abstracted measures of trade dependence are in the spirit of my theory or question.
+I hope to provide a more specific measure of trade dependence.
+I select the maximum value for theoretical reasons.
+Happily, I find a statistically significant relationship that I can explore.
 
@@
 Let's look at the relationship between the predicted probability that state $i$ will initiate a MID with state $j$ for a given level of product-level trade dependence.
-First, I will explore the different degrees of effects predicted using the different measures of product-level trade dependence.
-I will then compare the relationship between MID initiation and product-level and broad, total trade dependence.
-Finally, I will include traditional controls for a state's decision to initiate a MID against a rival to see whether trade dependence is a significant factor in this complex decision.
+First, I will compare the relationship between MID initiation and product-level and broad, total trade dependence.
+I will then include traditional controls for a state's decision to initiate a MID against a rival to see whether trade dependence is a significant factor in this complex decision.
 

### 2022-12-12 analysis/models.R

@@ -6,12 +6,66 @@ library(labelled)
 # Load in full dataset ----------------------------------------------------
 
 perc_df <- rio::import(here::here("data", "full_df.csv")) |> 
-  mutate(across(trade_dep_total:trade_dep_broad, ~ .x * 100)) |> 
+  mutate(across(trade_dep_max:trade_dep_broad, ~ .x * 100),
+         trade_dep_total = trade_dep_max) |> 
   set_variable_labels(trade_dep_total = "Trade dependence, product-level",
                       trade_dep_broad = "Trade dependence, total trade", 
                       conttype = "Contiguous",
                       v2x_polyarchy1 = "Democracy, state i",
                       v2x_polyarchy2 = "Democracy, state j")
+
+# Product-level trade dependence measures ---------------------------------
+
+m_max <- glm(
+  mid_onset_all ~ trade_dep_max, 
+  family = binomial(link = "logit"), 
+  data = perc_df
+)
+
+m_75 <- glm(
+  mid_onset_all ~ trade_dep_75, 
+  family = binomial(link = "logit"), 
+  data = perc_df
+)
+
+m_avg <- glm(
+  mid_onset_all ~ trade_dep_avg, 
+  family = binomial(link = "logit"), 
+  data = perc_df
+)
+
+m_25 <- glm(
+  mid_onset_all ~ trade_dep_25, 
+  family = binomial(link = "logit"), 
+  data = perc_df
+)
+
+tidy(m_max, exponentiate = T) |> 
+  select(term, estimate, std.error) |> 
+  bind_rows(
+    tidy(m_75, exponentiate = T) |> 
+      select(term, estimate, std.error)
+  ) |> 
+  bind_rows(
+    tidy(m_avg, exponentiate = T) |> 
+      select(term, estimate, std.error)
+  ) |> 
+  bind_rows(
+    tidy(m_25, exponentiate = T) |> 
+      select(term, estimate, std.error)
+  ) |> 
+  filter(term != "(Intercept)") |> 
+  mutate(
+    conf.low = estimate - (std.error * 1.96),
+    conf.high = estimate + (std.error * 1.96),
+    term = factor(term, ordered = T)
+  ) |> 
+  select(-std.error) |> 
+  pivot_longer(estimate:conf.high) |> 
+  ggplot(aes(x = value, y = reorder(term, term))) +
+  geom_line() + 
+  geom_point() + 
+  geom_vline(xintercept = 0, colour = "darkgrey")
 
 # Basic models ------------------------------------------------------------
 

### 2022-12-12 data-processing/data-processing.R

@@ -72,9 +72,34 @@ trade_dep <- trade_bilat |>
 
 trade_dep_max <- trade_dep |> 
   group_by(country1, country2, year) |> 
+  filter(trade_dep_total > 0) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
   select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
+
+trade_dep_75 <- trade_dep |> 
+  group_by(country1, country2, year) |>
+  filter(trade_dep_total > 0) |> 
+  summarise(trade_dep_75 = quantile(trade_dep_total)[[4]]) |> 
+  ungroup()
+
+trade_dep_avg <- trade_dep |> 
+  group_by(country1, country2, year) |> 
+  filter(trade_dep_total > 0) |> 
+  summarise(trade_dep_avg = mean(trade_dep_total)) |> 
+  ungroup()
+
+trade_dep_25 <- trade_dep |> 
+  group_by(country1, country2, year) |> 
+  filter(trade_dep_total > 0) |> 
+  summarise(trade_dep_25 = quantile(trade_dep_total)[[2]]) |> 
+  ungroup()
+
+trade_dep_min <- trade_dep |> 
+  group_by(country1, country2, year) |> 
+  filter(trade_dep_total > 0) |> 
+  summarise(trade_dep_min = min(trade_dep_total)) |> 
+  ungroup()
 
 trade_dep_broad <- trade_bilat |> 
   group_by(year, country1, country2) |> 
@@ -105,9 +130,14 @@ controls <- scope |>
 # TODO: Work out what are true NAs and what are 0s. 
 full_df <- mid_df |> 
   left_join(trade_dep_max, by = c("country1", "country2", "year")) |> 
+  left_join(trade_dep_75, by = c("country1", "country2", "year")) |> 
+  left_join(trade_dep_avg, by = c("country1", "country2", "year")) |> 
+  left_join(trade_dep_25, by = c("country1", "country2", "year")) |> 
+  left_join(trade_dep_min, by = c("country1", "country2", "year")) |> 
   left_join(trade_dep_broad, by = c("country1", "country2", "year")) |>
   left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year")) |> 
-  mutate(across(trade_dep_total:trade_dep_broad, ~ replace_na(.x, 0)))
[…]

### 2022-12-13 article/article.qmd
Update output to pdf

@@
 ---
-title: "Trade Dependence"
+title: "Trade Dependence and Militarized Interstate Disputes"
 format: 
-  html:
-    self-contained: true
+  pdf:
+    documentclass: article
+    papersize: letter
+    geometry:
+      - margin=1in
+    linestretch: 2
 bibliography: bibliography.bib
@@
         plot.title = element_text(face = "bold"),
-        legend.position = "bottom")
+        legend.position = "bottom",
+        text = element_text(family = "serif"))
 
@@
 Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy.
-Using a more disaggregated measure of trade dependence, I find that dependence does not have a significant effect on the likelihood of militarized disputes between states.
+Using a more disaggregated measure of trade dependence, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.
 
@@
 We are interested in understanding how trade dependence can deter conflict.
-Rarely (if ever) do states threaten to sever their entire trade relationship with another state, including trade dependent states.
+Rarely (if ever) do states threaten to sever their entire trade relationship with another state.
 Rather, we regularly see states threaten or enact trade bans on particular products.
-In fact, states often aim to threaten to, or actually, inflict the greatest harm on their rival whilst minimizing the costs they themselves will face.
-A broad understanding of trade dependence does not capture the real threat a potential aggressor faces or should expect to face.
+In fact, states often threaten to inflict the greatest harm on their rival whilst minimizing the costs they will face.
+A broad understanding of trade dependence does not capture the real threat a potential aggressor faces.
 A more accurate measure of this factor is product-level trade dependence.
@@
 Before we explore this dynamic in detail, let us step out in detail the process we seek better understand.
-First, tensions between two states build.
-This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an unfavourable bilateral foreign policy.
+First, tensions between two states builds.
+This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an controversial bilateral foreign policy.
 The threatened state issues a formal request to the initiating state to change or reverse their policy actions.
 The initiating state refuses: the incentives for enacting the controversial policy continue to outweigh the cost of a soured relationship with the threatened state.
-The threatened state then states that it will adopt punitive policies towards the initiator, including trade embargoes on specific products on which the initiator depends.
+The threatened state responds by threatening to adopt punitive policies towards the initiator, including trade embargoes on specific products on which the initiator depends.
 Here, the initiator's expected costs from pursuing this policy change: they know that if they continue along this path they will likely face economic costs in the form of trade disruptions.
@@
 This is a directed question.
-We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on their other.
+We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on the other.
 
@@
 
-Following other literature to enhance comparability, I include the following controls.
+Following other literature exploring this question, I include several control variables.
+I include a binary variable accounting for whether the dyad includes states that share a land border.
+I also include two variables providing each state's level of democracy.
 
@@
 
-We will compare this to the predicted probability derived using the traditional broad measure of trade dependence.
+First, let us compare the relationship between MID initiation and product-level trade dependence compared to the relationship estimated using the more traditional, broad measure of trade dependence.
 
 ```{r}
+#| tbl-cap: Base models
+
 tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
@@
 
-*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^3]
+*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^2]
 
-[^3]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.
+[^2]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.
 
 ```{r}
+#| fig-cap: Predicted probability of MID given product-level trade dependence
 ggplot(pred_m1_m2, aes(x = trade_dep, y = .fitted, colour = type)) + 
@@
   labs(
-    title = "Predicted probability of MID given product-level trade dependence",
     x = "Trade dependence (%)",
@@
 
-Moving from these base models to ones that incorporate traditional controls for MIDs, we learn that the relationship between conflict and trade dependence is not significant.
+Moving from these base models to ones that incorporate traditional controls for MIDs and trade relationships, we learn that the relationship between conflict and trade dependence is not significant.
 
 ```{r}
+#| tbl-cap: Controlled models
+
 tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
@@
 Significant determinants of conflict include contiguity and state $i$'s level of democracy.
-Concentrating on product-level trade dependence, I find that state $i$ is less to initiate an MID against a state with which it shares a land border.
+I find that state $i$ is less to initiate an MID against a state with which it shares a land border, on average holding all else constant.
 I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$.
-Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID, holding all else constant.
+Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID.
+
+{{< pagebreak >}}
+
+## Bibliography

### 2022-12-13 analysis/models.R

@@ -14,59 +14,6 @@ perc_df <- rio::import(here::here("data", "full_df.csv")) |>
                       v2x_polyarchy1 = "Democracy, state i",
                       v2x_polyarchy2 = "Democracy, state j")
 
-# Product-level trade dependence measures ---------------------------------
-
-m_max <- glm(
-  mid_onset_all ~ trade_dep_max, 
-  family = binomial(link = "logit"), 
-  data = perc_df
-)
-
-m_75 <- glm(
-  mid_onset_all ~ trade_dep_75, 
-  family = binomial(link = "logit"), 
-  data = perc_df
-)
-
-m_avg <- glm(
-  mid_onset_all ~ trade_dep_avg, 
-  family = binomial(link = "logit"), 
-  data = perc_df
-)
-
-m_25 <- glm(
-  mid_onset_all ~ trade_dep_25, 
-  family = binomial(link = "logit"), 
-  data = perc_df
-)
-
-tidy(m_max, exponentiate = T) |> 
-  select(term, estimate, std.error) |> 
-  bind_rows(
-    tidy(m_75, exponentiate = T) |> 
-      select(term, estimate, std.error)
-  ) |> 
-  bind_rows(
-    tidy(m_avg, exponentiate = T) |> 
-      select(term, estimate, std.error)
-  ) |> 
-  bind_rows(
-    tidy(m_25, exponentiate = T) |> 
-      select(term, estimate, std.error)
-  ) |> 
-  filter(term != "(Intercept)") |> 
-  mutate(
-    conf.low = estimate - (std.error * 1.96),
-    conf.high = estimate + (std.error * 1.96),
-    term = factor(term, ordered = T)
-  ) |> 
-  select(-std.error) |> 
-  pivot_longer(estimate:conf.high) |> 
-  ggplot(aes(x = value, y = reorder(term, term))) +
-  geom_line() + 
-  geom_point() + 
-  geom_vline(xintercept = 0, colour = "darkgrey")
-
 # Basic models ------------------------------------------------------------
 
 m1 <- glm(

### 2022-12-13 data-processing/data-processing.R

@@ -75,7 +75,8 @@ trade_dep_max <- trade_dep |>
   filter(trade_dep_total > 0) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
-  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
+  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total) |> 
+  mutate(year = 2015)
 
 trade_dep_75 <- trade_dep |> 
   group_by(country1, country2, year) |>
@@ -130,10 +131,6 @@ controls <- scope |>
 # TODO: Work out what are true NAs and what are 0s. 
 full_df <- mid_df |> 
   left_join(trade_dep_max, by = c("country1", "country2", "year")) |> 
-  left_join(trade_dep_75, by = c("country1", "country2", "year")) |> 
-  left_join(trade_dep_avg, by = c("country1", "country2", "year")) |> 
-  left_join(trade_dep_25, by = c("country1", "country2", "year")) |> 
-  left_join(trade_dep_min, by = c("country1", "country2", "year")) |> 
   left_join(trade_dep_broad, by = c("country1", "country2", "year")) |>
   left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year")) |> 
   mutate(across(trade_dep_max:trade_dep_broad, ~ replace_na(.x, 0)),

### 2022-12-13 src/src_trade.R

@@ -25,12 +25,6 @@ reporter_ids <- fromJSON("https://comtrade.un.org/Data/cache/reporterAreas.json"
                       "Fmr Ethiopia")) |> 
   pull(id)
 
-pull_scope <- tibble(
-  year = 2008:2013
-) |> 
-  full_join(tibble(reporter_id = reporter_ids), by = character()) |> 
-  arrange(desc(year))
-
 pull_comtrade <- function(reporter_id, year) {
   
   query <- glue::glue("https://comtrade.un.org/api/get?r={ reporter_id }&ps={ year }&fmt=json&freq=A&head=M&px=HS&cc=AG2")
@@ -52,7 +46,8 @@ pull_comtrade <- function(reporter_id, year) {
   if (result$validation$status$name != "Result too large" & length(result$dataset) > 0) {
     
     return(
-      tibble(result$dataset)
+      tibble(result$dataset) |> 
+        mutate(across(everything(), ~ as.character(.x)))
     )
     
   } else if (result$validation$status$name != "Result too large" & length(result$dataset) == 0) {
@@ -68,7 +63,7 @@ pull_comtrade <- function(reporter_id, year) {
     
     return(
       tibble(
-        rtCode = as.integer(reporter_id),
+        rtCode = reporter_id,
         status = "Result too large"
       )
     )
@@ -77,7 +72,7 @@ pull_comtrade <- function(reporter_id, year) {
   
 }
 
-trade_df <- map2_dfr(pull_scope$reporter_id, pull_scope$year, ~ pull_comtrade(.x, .y))
+safely_pull_comtrade <- safely(~ pull_comtrade(.x, .y))
 
-
-
+trade_df <- map2_dfr(reporter_ids, 2013, ~ pull_comtrade(.x, .y))
+ 

### 2022-12-13 article/article.qmd
Update artilce

@@
 library(gtsummary)
-library(modelsummary)
 ```
@@
 Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy.
-Using a more disaggregated measure of trade dependence, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.
+Using this more disaggregated measure of trade dependence and a binary logistic regression model, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.
 
@@
 Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation.
-Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error.
-Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP).
+Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error [@chen2021; @chatagnier2017].
+Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP) [@chen2021; @flores-macias2013b; @gartzke2016; @kinne2012a; @lupu2013; @pevehouse2004].
 However, a more disaggregated approach to trade dependence is worth exploring.
@@
 
-This analysis focuses on all directed state dyads in 2014.
+This analysis focuses on all directed state dyads in 2013 and 2014.
 This includes `r nrow(full_df) |> comma()` observations.
@@
 The narrow timeframe included in this analysis is driven by data constraints.
-I only have access to complete trade and militarized interstate dispute data for 2014.
+I only have access to complete trade and militarized interstate dispute data for 2013 and 2014.
 A natural first step for extending this analysis is including a broader timeframe in the analysis.
@@
 This dataset includes all known MIDs from 1816-2014, recorded at the participant level.
-In 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.
+Between 2013 and 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.
 
@@
 
-In 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
+In 2013 and 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
 The most dependent pairing is between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`: `r pull(trade_dep_max, country1)`'s trade in `r pull(trade_dep_max, cmd_desc_e)` with `r pull(trade_dep_max, country2)` amounts to `r trade_dep_max |> pull(trade_dep_total) |> percent()` of `r pull(trade_dep_max, country1)`'s GDP in 2014.
@@
 
-```{r}
-#| tbl-cap: Base models
-
-tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
-```
+**See Appendix for Table 1**
 
@@
 [^2]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.
+
+**See Appendix for Figure 1**
+
+It is worth noting that product-level trade dependence predicts a much more aggressive relationship between conflict and trade dependence: the point at which MIDs are predicted to be more likely than not is at `r pred_m1_m2 |> filter(type == "Product-level", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for product-level trade dependence and `r pred_m1_m2 |> filter(type == "Total trade", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for total trade dependence.
+
+Moving from these base models to ones that incorporate traditional controls for MIDs and trade relationships, we learn that the relationship between conflict and trade dependence is not significant.
+
+**See Appendix for Table 2**
+
+For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds.
+Significant determinants of conflict include contiguity and state $i$'s level of democracy.
+I find that state $i$ is less to initiate an MID against a state with which it shares a land border, on average holding all else constant.
+I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$.
+Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID.
+
+{{< pagebreak >}}
+
+## Appendix
+
+### Regression results
+
+```{r}
+#| tbl-cap: Base models
+
+tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
+```
+
+```{r}
+#| tbl-cap: Controlled models
+
+tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
+```
+
+### Comparison of product-level and total trade dependence results
 
@@
 
-It is worth noting that product-level trade dependence predicts a much more aggressive relationship between conflict and trade dependence: the point at which MIDs are predicted to be more likely than not is at `r pred_m1_m2 |> filter(type == "Product-level", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for product-level trade dependence and `r pred_m1_m2 |> filter(type == "Total trade", .fitted >= 0.5) |> slice(1) |> pull(trade_dep)`% for total trade dependence.
-
-Moving from these base models to ones that incorporate traditional controls for MIDs and trade relationships, we learn that the relationship between conflict and trade dependence is not significant.
-
-```{r}
-#| tbl-cap: Controlled models
-
-tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
-```
-
-For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds.
-Significant determinants of conflict include contiguity and state $i$'s level of democracy.
-I find that state $i$ is less to initiate an MID against a state with which it shares a land border, on average holding all else constant.
-I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$.
-Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID.
-
 {{< pagebreak >}}

### 2022-12-13 article/article.tex
Update artilce

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Trade Dependence and Militarized Interstate Disputes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Trade Dependence and Militarized Interstate Disputes}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

\setstretch{2}
Does trade dependence reduce conflict between states? Debate remains
over the role and significance of economic interdependence in promoting
peace: different scholars have found that dependence increases,
decreases, and has no significant effect on the likelihood of conflict
between states. However, this research has generally relied on a broad
measure of bilateral trade dependence. This paper will delve deeper into
the effect of this economic relationship by exploring product-level
dependence. If trade dependence deters conflict between states by
increasing the expected cost of conflict through trade disruptions, we
should measure the true cost of this disruption. Rarely, if ever, do we
see states threaten to or impose total trade embargoes. Rather, states
impose trade bans on specific products that maximize the cost borne by
their opponent while minimizing the cost to their own economy. Using
this more disaggregated measure of trade dependence and a binary
logistic regression model, I find that this dependence does not have a
significant effect on the likelihood of militarized disputes between
states.

\newpage{}

The commercial peace theory maintains that trade fosters peace between
states because it makes conflict more costly. When a state is deciding
whether to initiate conflict against another state, they weigh the
benefits they expect to derive against the costs they expect to incur.
When the benefits outweigh the costs, conflict begins. Commercial peace
theory is a simple extension of this bargaining theory of conflict: the
inevitable trade disruptions caused by conflict increase the costs
states face, which decreases the likelihood that the benefits will
outweigh the costs and conflict will begin, \emph{ceteris paribus}.

Debate remains over the role played by, and significance of, bilateral
trade dependence in conflict mitigation. Recent scholarship on the topic
has begun exploring whether this ambiguity is the product of measurement
error (Chen 2021; Chatagnier and Kavaklı 2017). Within international
relations scholarship, measures of trade dependence have been defined at
a broad level, predominantly using a measure introduced by Oneal and
Russett (1997) that explores total trade between two states relative to
their gross domestic product (GDP) (Chen 2021; Flores-Macías and Kreps
2013; Gartzke and Westerwinter 2016; Kinne 2012; Lupu and Traag 2013;
Pevehouse 2004). However, a more disaggregated approach to trade
dependence is worth exploring. We are interested in understanding how
trade dependence can deter conflict. Rarely (if ever) do states threaten
to sever their entire trade relationship with another state. Rather, we
regularly see states threaten or enact trade bans on particular
products. In fact, states often threaten to inflict the greatest harm on
their rival whilst minimizing the costs they will face. A broad
understanding of trade dependence does not capture the real threat a
potential aggressor faces. A more accurate measure of this factor is
product-level trade dependence. Asymmetries at this level have been used
as a foreign policy mechanism to deter or punish rival states.

Before we explore this dynamic in detail, let us step out in detail the
process we seek better understand. First, tensions between two states
builds. This could be for a number of reasons, including historical
animosity, a diplomatic spat over a regional or global issue, or
domestic political interests prompting an controversial bilateral
foreign policy. The threatened state issues a formal request to the
initiating state to change or reverse their policy actions. The
initiating state refuses: the incentives for enacting the controversial
policy continue to outweigh the cost of a soured relationship with the
threatened state. The threatened state responds by threatening to adopt
punitive policies towards the initiator, including trade embargoes on
specific products on which the initiator depends. Here, the initiator's
expected costs from pursuing this policy change: they know that if they
continue along this path they will likely face economic costs in the
form of trade disruptions. Does this deter them? Many scholars would
ask: how costly is this disruption? This paper will answer this
question. If the costs are not so great, the initiator may respond to
this threat through further escalation, including through a militarized
dispute. International relations scholarship uses this broad measure of
conflict to include actions that sit below invasion or war. This is
often the precursor to full-scale interstate war. States use these
actions as an extreme signal of their intent to escalate a dispute to
war, hoping to deter their opponent.

We are interested in the relationship between trade dependence and this
pre-cursor to war. Certainly, if two states are at war with one another,
we should expect trade to nearly, if not entirely, cease between them.
However, we should not expect trade to entirely cease between states
when they are still circling each other, hoping to achieve their most
desired policy outcome and avoid war. Tying a measure of the entire
trade relationship to whether or not states initiate militarized
interstate disputes (not war) against each other misses this important
distinction. This paper seeks to address this.

\hypertarget{trade-dependence-and-militarized-conflict}{%
\subsection{Trade dependence and militarized
conflict}\label{trade-dependence-and-militarized-conflict}}

Returning to our motivating question: does trade dependence deter states
from initiating militarized conflict against each other? The commercial
peace theory argues that trade dependence increases the expected cost of
initiating conflict against a trade partner, decreasing the expected net
benefits derived from such an action and, therefore, reducing the
likelihood a state will initiate conflict. This leads to the following
testable hypothesis:

\begin{quote}
H1: As product-level trade dependence between two states increases, the
likelihood that a state will initiate militarized conflict against its
trade partner decreases.
\end{quote}

\hypertarget{research-design}{%
\subsection{Research design}\label{research-design}}

\hypertarget{scope}{%
\subsubsection{Scope}\label{scope}}

This analysis focuses on all directed state dyads in 2013 and 2014. This
includes 74,884 observations. We are interested in whether or not a
trade dependent state initiates conflict against another state. This is
a directed question. We need to know whether the potential initiator is
dependent on their opponent, not simply whether either partner has a
dependence on the other.

The narrow timeframe included in this analysis is driven by data
constraints. I only have access to complete trade and militarized
interstate dispute data for 2013 and 2014. A natural first step for
extending this analysis is including a broader timeframe in the
analysis.

\hypertarget{dependent-variable}{%
\subsubsection{Dependent variable}\label{dependent-variable}}

My dependent variable focuses on the onset of militarized interstate
disputes (MIDs), or events in which one or more states threaten,
display, or use force against another state or states. If state \(i\)
initiated the MID against state \(j\) in a given year, this variable is
coded 1. Otherwise, it receives a value of 0.

The data come from version five of Palmer et al's (2022) Militarized
Interstate Dispute data collection. This dataset includes all known MIDs
from 1816-2014, recorded at the participant level. Between 2013 and
2014, 135 MIDs were initiated.

\hypertarget{independent-variables-of-interest}{%
\subsubsection{Independent variables of
interest}\label{independent-variables-of-interest}}

My primary independent variable of interest is bilateral trade
dependence. This paper introduces a novel measure of this dependence.
Rather than looking at the balance of trade in total imports, exports,
or trade value, I measure dependence in trade of specific products. This
more fine-grained approach to measuring trade dependence allows us to
examine more precisely the potential costs an initiator expects to incur
if they escalate a dispute to a MID.

To do this, I adopt the classical structure proposed by Oneal and
Russett (1997) but adapt it to trade in each product. I use bilateral
annual export and import data provided by UN Comtrade. Commodities are
grouped according to the standard two digit harmonized system. For
example, all live animals, including horses, bovines, swine, sheep, and
goats, are grouped together into a category called \emph{animals -
live}. My theory assumes a degree of substitution within this group. It
would be interesting to explore the level of granularity required to
meaningfully make this assumption. Sadly, this is difficult to test
because we lose data completeness as we increase granularity.

Formally, my measure of trade dependence is defined as the total imports
and exports from state \(i\) to (or from) state \(j\) for a given
commodity in a given year relative to state \(i\)'s GDP for that year.

\[
r_{ij,p,t} = \frac{m_{ij,p,t} + x_{ij,p,t}}{GDP_{i,t}}
\]

Where \(j\) is the trade partner and \(p\) is the product. This value is
measured for each year, \(t\).

In 2013 and 2014, product-level trade dependence ranged from 0\% to
36\%, with an average level of trade dependence of 0.09\%. The most
dependent pairing is between Sudan and United Arab Emirates: Sudan's
trade in natural, cultured pearls; precious, semi-precious stones;
precious metals, metals clad with precious metal, and articles thereof;
imitation jewellery; coin with United Arab Emirates amounts to 36\% of
Sudan's GDP in 2014.

States trade in a large number of commodities. For each directed state
dyad, we get the same number of measures of dependence as products
traded between them. We need to reduce this to a single measure of
dependence. I use the product with the highest value of dependence
between states \(i\) and \(j\). State \(j\) will strategically select
the product or products that maximize the cost state \(i\) should expect
to incur in retaliation for escalating tensions. The trade in a product
on which state \(i\) is most dependent is a good candidate for this
strategic threat or imposition of a trade embargo.\footnote{I tested
  five options for this aggregation: the minimum, the 25th percentile,
  the average, the 75th percentile, and the maximum level of trade
  dependence between two states. Let's first discuss the logic of the
  extremes: the minimum and maximum values. The minimum value represents
  the lowest possible cost state \(j\) could threaten to impose on state
  \(i\) in a given year. If this is relatively high, state \(i\) is
  probably very trade dependent on state \(j\). This should be a very
  strong measure of dependence. However, state \(j\) is unlikely to
  threaten to stop trade in this commodity. Remember, they aim to impose
  the greatest possible cost on state \(i\) whilst minimizing the cost
  to themselves. Perhaps a better candidate is trade in the most
  dependent commodity. To qualify these measures, I also also tested
  intermediate measures of dependence: the 25th, 50th, and 75th
  percentiles of trade dependence across all traded commodities. These
  measures perhaps better capture the dynamics at play with the extreme
  measures whilst allowing state \(j\) to minimize the cost it bears and
  also to select a number of commodities to disrupt, not just one. I
  found that the maximum, 75th percentile, and average measures of trade
  dependence were statistically significant to initiation of MIDs. In
  fact, the average measure of trade dependence produced the strongest
  effect on MID initiation. However, I am not sure that these abstracted
  measures of trade dependence are in the spirit of my theory or
  question. I hope to provide a more specific measure of trade
  dependence. I select the maximum value for theoretical reasons.
  Happily, I find a statistically significant relationship that I can
  explore.}

This measure allows me to focus on trade in goods that are critical to
the potential initiator. I argue that this better represents the true
cost they expect to incur if they escalate tensions with a trade partner
to a MID.

\hypertarget{control-variables}{%
\subsubsection{Control variables}\label{control-variables}}

Following other literature exploring this question, I include several
control variables. I include a binary variable accounting for whether
the dyad includes states that share a land border. I also include two
variables providing each state's level of democracy.

\hypertarget{results}{%
\subsection{Results}\label{results}}

Let's look at the relationship between the predicted probability that
state \(i\) will initiate a MID with state \(j\) for a given level of
product-level trade dependence. First, I will compare the relationship
between MID initiation and product-level and broad, total trade
dependence. I will then include traditional controls for a state's
decision to initiate a MID against a rival to see whether trade
dependence is a significant factor in this complex decision.

First, let us compare the relationship between MID initiation and
product-level trade dependence compared to the relationship estimated
using the more traditional, broad measure of trade dependence.

\textbf{See Appendix for Table 1}

I used a binary logistic regression model to predict the probability
that state \(i\) initiates a MID against state \(j\) for a given level
of trade dependence. Using product-level trade dependence, I find that
increasing dependence by one percentage point \textbf{increases} the
odds that state \(i\) will initiate MID against state \(j\) by 1.19 on
average. This relationship is highly statistically significant. This
more fine-grained measure of trade dependence predicts a stronger
relationship between conflict and trade dependence than the more
tradition and broad-based approach. Using total trade dependence, I find
that increasing dependence by one percentage point increases the odds
that state \(i\) will initiate MID against state \(j\) by 1.09 on
average.

\emph{Figure 1} illustrates the different predicted probabilities that
state \(i\) will initiate an MID against state \(j\) for each possible
level trade dependence between no dependence (0 percent) and total
dependence (100 percent).\footnote{Any prediction made for a level of
  trade dependence above 36\% for product-level dependence and 71\% for
  total trade dependence is an out of sample prediction.}

\textbf{See Appendix for Figure 1}

It is worth noting that product-level trade dependence predicts a much
more aggressive relationship between conflict and trade dependence: the
point at which MIDs are predicted to be more likely than not is at 36\%
for product-level trade dependence and 71\% for total trade dependence.

Moving from these base models to ones that incorporate traditional
controls for MIDs and trade relationships, we learn that the
relationship between conflict and trade dependence is not significant.

\textbf{See Appendix for Table 2}

For both product-level and total trade dependence, the odds that state
\(i\) will initiate an MID against state \(j\) remain positive; however,
the statistical significance of this relationship drops well below
traditional thresholds. Significant determinants of conflict include
contiguity and state \(i\)'s level of democracy. I find that state \(i\)
is less to initiate an MID against a state with which it shares a land
border, on average holding all else constant. I also find that as state
\(i\) becomes more democratic, it is less likely to initiate a MID
against state \(j\). Interestingly, I find that state \(j\)'s level of
democracy does not appear to be a significant factor in state \(i\)'s
decision to initiate an MID.

\newpage{}

\hypertarget{appendix}{%
\subsection{Appendix}\label{appendix}}

\hypertarget{regression-results}{%
\subsubsection{Regression results}\label{regression-results}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3265}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0816}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1224}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1327}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0816}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1224}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1327}}@{}}
\caption{Base models }\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Characteristic}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Characteristic}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} \\
\midrule()
\endhead
(Intercept) & 0.00 & 0.00, 0.00 & \textless0.001 & 0.00 & 0.00, 0.00 &
\textless0.001 \\
Trade dependence, product-level & 1.19 & 1.13, 1.25 & \textless0.001 & &
& \\
Trade dependence, total trade & & & & 1.09 & 1.07, 1.12 &
\textless0.001 \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3265}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0816}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1224}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1327}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0816}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1224}}
  >{\centering\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1327}}@{}}
\caption{Controlled models }\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Characteristic}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Characteristic}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{OR}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{95\% CI}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{p-value}
\end{minipage} \\
\midrule()
\endhead
(Intercept) & 0.00 & 0.00, 0.00 & \textless0.001 & 0.00 & 0.00, 0.00 &
\textless0.001 \\
Trade dependence, product-level & 1.05 & 0.96, 1.12 & 0.2 & & & \\
Contiguous & 69.7 & 48.6, 101 & \textless0.001 & 71.3 & 49.5, 103 &
\textless0.001 \\
Democracy, state i & 0.13 & 0.06, 0.27 & \textless0.001 & 0.13 & 0.06,
0.27 & \textless0.001 \\
Democracy, state j & 1.02 & 0.49, 2.11 & \textgreater0.9 & 1.00 & 0.48,
2.08 & \textgreater0.9 \\
Trade dependence, total trade & & & & 1.00 & 0.97, 1.03 & 0.8 \\
\bottomrule()
\end{longtable}

\hypertarget{comparison-of-product-level-and-total-trade-dependence-results}{%
\subsubsection{Comparison of product-level and total trade dependence
results}\label{comparison-of-product-level-and-total-trade-dependence-results}}

\begin{figure}

{\centering \includegraphics{article_files/figure-pdf/unnamed-chunk-4-1.pdf}

}

\caption{Predicted probability of MID given product-level trade
dependence}

\end{figure}

\newpage{}

\hypertarget{bibliography}{%
\subsection*{Bibliography}\label{bibliography}}
\addcontentsline{toc}{subsection}{Bibliography}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-chatagnier2017}{}}%
Chatagnier, J. Tyson, and Kerim Can Kavaklı. 2017. {``From {Economic
Competition} to {Military Combat}: {Export Similarity} and
{International Conflict}.''} \emph{The Journal of Conflict Resolution}
61 (7): 1510--36. \url{https://www.jstor.org/stable/26363938}.

\leavevmode\vadjust pre{\hypertarget{ref-chen2021}{}}%
Chen, Frederick R. 2021. {``Extended {Dependence}: {Trade}, {Alliances},
and {Peace}.''} \emph{The Journal of Politics} 83 (1): 246--59.
\url{https://doi.org/10.1086/709149}.

\leavevmode\vadjust pre{\hypertarget{ref-flores-macias2013b}{}}%
Flores-Macías, Gustavo A., and Sarah E. Kreps. 2013. {``The {Foreign
Policy Consequences} of {Trade}: {China}'s {Commercial Relations} with
{Africa} and {Latin America}, 1992--2006.''} \emph{The Journal of
Politics} 75 (2): 357--71.
\url{https://doi.org/10.1017/S0022381613000066}.

\leavevmode\vadjust pre{\hypertarget{ref-gartzke2016}{}}%
Gartzke, Erik, and Oliver Westerwinter. 2016. {``The Complex Structure
of Commercial Peace Contrasting Trade Interdependence, Asymmetry, and
Multipolarity.''} \emph{Journal of Peace Research} 53 (3): 325--43.
\url{https://www.jstor.org/stable/43920593}.

\leavevmode\vadjust pre{\hypertarget{ref-kinne2012a}{}}%
Kinne, Brandon J. 2012. {``Multilateral {Trade} and {Militarized
Conflict}: {Centrality}, {Openness}, and {Asymmetry} in the {Global
Trade Network}.''} \emph{The Journal of Politics} 74 (1): 308--22.
\url{https://doi.org/10.1017/S002238161100137X}.

\leavevmode\vadjust pre{\hypertarget{ref-lupu2013}{}}%
Lupu, Yonatan, and Vincent A. Traag. 2013. {``Trading {Communities}, the
{Networked Structure} of {International Relations}, and the {Kantian
Peace}.''} \emph{The Journal of Conflict Resolution} 57 (6): 1011--42.
\url{https://www.jstor.org/stable/24545601}.

\leavevmode\vadjust pre{\hypertarget{ref-oneal1997}{}}%
Oneal, John R., and Bruce M. Russet. 1997. {``The Classical Liberals
Were Right: Democracy, Interdependence, and Conflict, 1950-1985.''}
\emph{International Studies Quarterly} 41 (2): 267--94.
\url{https://doi.org/10.1111/1468-2478.00042}.

\leavevmode\vadjust pre{\hypertarget{ref-palmer2022}{}}%
Palmer, Glenn, Roseanne W McManus, Vito D'Orazio, Michael R Kenwick,
Mikaela Karstens, Chase Bloch, Nick Dietrich, Kayla Kahn, Kellan Ritter,
and Michael J Soules. 2022. {``The {Mid5 Dataset}, 2011--2014:
{Procedures}, Coding Rules, and Description.''} \emph{Conflict
Management and Peace Science} 39 (4): 470--82.
\url{https://doi.org/10.1177/0738894221995743}.

\leavevmode\vadjust pre{\hypertarget{ref-pevehouse2004}{}}%
Pevehouse, Jon C. 2004. {``Interdependence {Theory} and the
{Measurement} of {International Conflict}.''} \emph{The Journal of
Politics} 66 (1): 247--66.
\url{https://doi.org/10.1046/j.1468-2508.2004.00150.x}.

\end{CSLReferences}

\end{document}

### 2022-12-13 data-processing/data-processing.R
Update trade data

@@ -72,35 +72,16 @@ trade_dep <- trade_bilat |>
 
 trade_dep_max <- trade_dep |> 
   group_by(country1, country2, year) |> 
-  filter(trade_dep_total > 0) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
-  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total) |> 
-  mutate(year = 2015)
+  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
 
 trade_dep_75 <- trade_dep |> 
-  group_by(country1, country2, year) |>
-  filter(trade_dep_total > 0) |> 
-  summarise(trade_dep_75 = quantile(trade_dep_total)[[4]]) |> 
-  ungroup()
-
-trade_dep_avg <- trade_dep |> 
   group_by(country1, country2, year) |> 
-  filter(trade_dep_total > 0) |> 
-  summarise(trade_dep_avg = mean(trade_dep_total)) |> 
-  ungroup()
-
-trade_dep_25 <- trade_dep |> 
-  group_by(country1, country2, year) |> 
-  filter(trade_dep_total > 0) |> 
-  summarise(trade_dep_25 = quantile(trade_dep_total)[[2]]) |> 
-  ungroup()
-
-trade_dep_min <- trade_dep |> 
-  group_by(country1, country2, year) |> 
-  filter(trade_dep_total > 0) |> 
-  summarise(trade_dep_min = min(trade_dep_total)) |> 
-  ungroup()
+  mutate(trade_dep_75 = quantile(trade_dep_total)[[4]]) |> 
+  view()
+  ungroup() |> 
+  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
 
 trade_dep_broad <- trade_bilat |> 
   group_by(year, country1, country2) |> 
@@ -133,8 +114,7 @@ full_df <- mid_df |>
   left_join(trade_dep_max, by = c("country1", "country2", "year")) |> 
   left_join(trade_dep_broad, by = c("country1", "country2", "year")) |>
   left_join(controls, by = c("ccode1", "country1", "ccode2", "country2", "year")) |> 
-  mutate(across(trade_dep_max:trade_dep_broad, ~ replace_na(.x, 0)),
-         trade_dep_total = trade_dep_max)
[…]

### 2022-12-13 src/src_trade.R

@@ -75,4 +75,6 @@ pull_comtrade <- function(reporter_id, year) {
 safely_pull_comtrade <- safely(~ pull_comtrade(.x, .y))
 
 trade_df <- map2_dfr(reporter_ids, 2013, ~ pull_comtrade(.x, .y))
+
+rio::export(trade_df, here::here("data-raw", "trade_data_2013.csv"))
  

### 2022-12-13 src/src_trade_large_countries.R

@@ -3,23 +3,27 @@ library(countrycode)
 library(httr)
 library(jsonlite)
 
+# Set file path
+year <- 2013
+
 # Read in existing trade data
-annual_trade_df <- rio::import(here::here("data-raw", "trade_data_2014.csv"))
+trade_df <- rio::import(here::here("data-raw", glue::glue("trade_data_{ year }.csv")))
 
 # Attempt direction-based pull --------------------------------------------
 
 # Find countries for which yearly trade data is too big
 pull_scope <- tibble(direction = c(1, 2)) |> 
   full_join(
-    rio::import(here::here("data-raw", "trade_data_2014.csv")) |> 
+    trade_df |> 
       filter(status == "Result too large") |> 
       select(rtCode),
     by = character()
-  )
+  ) |> 
+  mutate(across(everything(), ~ as.character(.x)))
 
 pull_large_comtrade <- function(reporter_id, direction) {
   
-  query <- glue::glue("https://comtrade.un.org/api/get?r={ reporter_id }&ps=2014&rg={ direction }&fmt=json&freq=A&head=M&px=HS&cc=AG2")
+  query <- glue::glue("https://comtrade.un.org/api/get?r={ reporter_id }&ps={ year }&rg={ direction }&fmt=json&freq=A&head=M&px=HS&cc=AG2")
   
   result <- tryCatch(
     fromJSON(query),
@@ -38,7 +42,8 @@ pull_large_comtrade <- function(reporter_id, direction) {
   if (result$validation$status$name != "Result too large" & length(result$dataset) > 0) {
     
     return(
-      tibble(result$dataset)
+      tibble(result$dataset) |> 
+        mutate(across(everything(), ~ as.character(.x)))
     )
     
   } else if (result$validation$status$name != "Result too large" & length(result$dataset) == 0) {
@@ -54,8 +59,8 @@ pull_large_comtrade <- function(reporter_id, direction) {
     
     return(
       tibble(
-        rtCode = as.integer(reporter_id),
-        rgCode = as.integer(direction),
+        rtCode = reporter_id,
+        rgCode = direction,
         status = glue::glue("Results too large: { direction }")
       )
     )
@@ -73,33 +78,37 @@ large_trade_df <- map2_dfr(
 )
 
 # Update results
-large_annual_trade_df <- annual_trade_df |> 
-  as_tibble() |> 
[…]

### 2022-12-13 article/article.qmd
Complete report

@@
 full_df <- rio::import(here::here("data", "full_df.csv"))
+sub_results <- rio::import(here::here("data", "sim_results.csv"))
 ```
@@
 This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence.
-If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption.
+If states' decision to escalate tensions with rivals depends on their expected cost from trade disruptions, we should measure the true cost of this disruption.
 Rarely, if ever, do we see states threaten to or impose total trade embargoes.
@@
 
-Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation.
+Another approach to this question, introduced by Barbieri [-@barbieri1996], argues that higher levels of trade dependence increase the likelihood that a dyad will engage in conflict.
+Peace between dyads is most likely to occur between equal partners: those states that are mutually dependent on trade with one another.
+As the dyad become less equal, both the more dominant and less dominant partner have incentives to weaken their trade partner through conflict.
+
+This debate over the role played by, and significance of, bilateral trade dependence in conflict mitigation remains unsettled.
 Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error [@chen2021; @chatagnier2017].
@@
 Rather, we regularly see states threaten or enact trade bans on particular products.
-In fact, states often threaten to inflict the greatest harm on their rival whilst minimizing the costs they will face.
+In fact, states aim to inflict the greatest harm on their rival whilst minimizing the costs they will face.
 A broad understanding of trade dependence does not capture the real threat a potential aggressor faces.
@@
 
-Before we explore this dynamic in detail, let us step out in detail the process we seek better understand.
-First, tensions between two states builds.
-This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an controversial bilateral foreign policy.
+Before we explore this dynamic, let us step out in detail the process we seek better understand.
+First, tensions between two states build.
+This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting a controversial bilateral foreign policy.
 The threatened state issues a formal request to the initiating state to change or reverse their policy actions.
@@
 This paper will answer this question.
-If the costs are not so great, the initiator may respond to this threat through further escalation, including through a militarized dispute.
+If the costs are great, the initiator may attempt to deescalate tensions.
+Alternatively, they may accept the short-term pain of trade disruptions to reduce their long-term dependence on their rival trade partner.
+What do we observe?
+
+We will look at whether the initiator state escalate the conflict to a militarized interstate dispute.
 International relations scholarship uses this broad measure of conflict to include actions that sit below invasion or war.
@@
 States use these actions as an extreme signal of their intent to escalate a dispute to war, hoping to deter their opponent.
-
 We are interested in the relationship between trade dependence and this pre-cursor to war.
@@
 However, we should not expect trade to entirely cease between states when they are still circling each other, hoping to achieve their most desired policy outcome and avoid war.
-Tying a measure of the entire trade relationship to whether or not states initiate militarized interstate disputes (not war) against each other misses this important distinction.
+Tying a measure of the entire trade relationship to whether or not states initiate militarized interstate disputes against each other misses this important distinction.
 This paper seeks to address this.
@@
 
+The alternative hypothesis argues that increasing trade dependence by one state on another leads to increased incentives for conflict. This leads to an alternative testable hypothesis:
+
+> H2: As product-level trade dependence between two states increases, the likelihood that a state will initiate militarized conflict against its trade partner increases.
+
 ## Research design
@@
 
-This analysis focuses on all directed state dyads in 2013 and 2014.
+This analysis focuses on all directed state dyads in 2014.
 This includes `r nrow(full_df) |> comma()` observations.
@@
 The narrow timeframe included in this analysis is driven by data constraints.
-I only have access to complete trade and militarized interstate dispute data for 2013 and 2014.
-A natural first step for extending this analysis is including a broader timeframe in the analysis.
+I only have access to complete trade and militarized interstate dispute data for 2014.
+A natural first step for extending this analysis is including a broader timeframe.
 
@@
 My dependent variable focuses on the onset of militarized interstate disputes (MIDs), or events in which one or more states threaten, display, or use force against another state or states.
-If state $i$ initiated the MID against state $j$ in a given year, this variable is coded 1.
+If state $i$ initiated a MID against state $j$ in a given year, this variable is coded 1.
 Otherwise, it receives a value of 0.
@@
 This dataset includes all known MIDs from 1816-2014, recorded at the participant level.
-Between 2013 and 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.
+In 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.
 
@@
 
-In 2013 and 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
+In 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
 The most dependent pairing is between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`: `r pull(trade_dep_max, country1)`'s trade in `r pull(trade_dep_max, cmd_desc_e)` with `r pull(trade_dep_max, country2)` amounts to `r trade_dep_max |> pull(trade_dep_total) |> percent()` of `r pull(trade_dep_max, country1)`'s GDP in 2014.
@@
 This relationship is highly statistically significant.
-This more fine-grained measure of trade dependence predicts a stronger relationship between conflict and trade dependence than the more tradition and broad-based approach.
-Using total trade dependence, I find that increasing dependence by one percentage point increases the odds that state $i$ will initiate MID against state $j$ by `r tidy(m2, exponentiate = T) |> filter(term == "trade_dep_broad") |> pull(estimate) |> round(2)` on average.
+This more fine-grained measure of trade dependence predicts a stronger relationship between conflict and trade dependence than the more traditional and broad-based approach.
+Using total trade dependence, I find that increasing dependence by one percentage point increases the odds that state $i$ will initiate a MID against state $j$ by `r tidy(m2, exponentiate = T) |> filter(term == "trade_dep_broad") |> pull(estimate) |> round(2)` on average.
 
-*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^2]
+*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate a MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^2]
 
@@
 
-For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds.
+For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain greater than one; however, the statistical significance of this relationship drops well below traditional thresholds.
 Significant determinants of conflict include contiguity and state $i$'s level of democracy.
-I find that state $i$ is less to initiate an MID against a state with which it shares a land border, on average holding all else constant.
+I find that state $i$ is less likely to initiate a MID against a state with which it shares a land border, on average holding all else constant.
 I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$.
-Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID.
+Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate a MID.
+
+Focusing on product-level trade dependence, let us explore the effect of moving from no trade dependence to the average product-level trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`.
+Following Hanmer and Kalkan [-@hanmer2013], I calculate the effect of this change setting all other variables to their observed values.
+I continue to find that moving from no trade dependence to the average level of trade dependence does not change the likelihood that state $i$ will initiate a MID against state $j$.
+However, this result is not statistically significant.
+
+What does this mean for policymakers and conflict scholars?
+First, I suspect that these results suffer from three empirical issues.
+First, the lack of complete bilateral, product-level trade data over time represents a major hurdle to conducting this analysis.
+I will continue to explore options to efficiently access these data.
+Second, my measure of trade dependence continues to not account for the networked nature of global trade markets.
+I am developing a more complex model of product-level trade dependence that accounts for indirect linkages (as advocated by Chen [-@chen2021]) and alternative partners.
+Without accounting for the whole network in which two states operate, we miss important factors in their calculations of their vulnerability to one another.
+Finally, my measure of trade dependence does not sufficiently account for the cost state $j$ faces in disrupting trade in state \$i\$'s most dependent product.
+State $j$ should attempt to maximize the cost to state $i$ of initiating a MID whilst minimizing the cost it will face.
+Rather than simply using the product that will impose the greatest harm on state $i$ in my analysis, I need to include a measure of the optimal choice for state $j$.
+I am working to incorporate this in my more complex model.
+
+In terms of substantive effects, I find that trade dependence does not matter to a state's decision to escalate conflict against a rival.
+If true, this is an important finding.
+Scholars and policymakers make many assumptions about a state's level of vulnerability on the global stage when deciding important foreign policy, including controversial foreign policy.
+Various power dynamics constrain or promote bold policy action.
+If states hope to reduce their vulnerability to conflict, they should not focus on developing a rival's economic dependence on them.
 

### 2022-12-13 analysis/models.R

@@ -2,12 +2,14 @@ library(tidyverse)
 library(broom)
 library(gtsummary)
 library(labelled)
+library(mvtnorm)
 
 # Load in full dataset ----------------------------------------------------
 
-perc_df <- rio::import(here::here("data", "full_df.csv")) |> 
-  mutate(across(trade_dep_max:trade_dep_broad, ~ .x * 100),
-         trade_dep_total = trade_dep_max) |> 
+full_df <- rio::import(here::here("data", "full_df.csv"))
+
+perc_df <- full_df |> 
+  mutate(across(trade_dep_total:trade_dep_broad, ~ .x * 100)) |> 
   set_variable_labels(trade_dep_total = "Trade dependence, product-level",
                       trade_dep_broad = "Trade dependence, total trade", 
                       conttype = "Contiguous",
@@ -48,7 +50,7 @@ pred_m1_m2 <- augment(
 # Controlled models -------------------------------------------------------
 
 m3 <- glm(
-  mid_onset_all ~ trade_dep_total + conttype + v2x_polyarchy1 + v2x_polyarchy2 , 
+  mid_onset_all ~ trade_dep_total + conttype + v2x_polyarchy1 + v2x_polyarchy2, 
   family = binomial(link = "logit"), 
   data = perc_df
 )

### 2022-12-13 analysis/substantive_effects.R

@@ -1,0 +1,60 @@
+library(mvtnorm)
+library(tidyverse)
+library(broom)
+
+# Substantive effects -----------------------------------------------------
+
+result_0 <- augment(m3, newdata = mutate(perc_df, trade_dep_total = 0), type.predict = "response") 
+result_0
+
+result_avg <- augment(m3, newdata = mutate(perc_df, trade_dep_total = mean(full_df$trade_dep_total) * 100), type.predict = "response") 
+result_avg
+
+result_diff <- result_0 |> 
+  bind_rows(result_avg) |> 
+  group_by(trade_dep_total) |> 
+  summarise(.fitted = mean(.fitted, na.rm = T)) |> 
+  mutate(diff = .fitted - lag(.fitted))
+result_diff
+
+coefs <- tidy(m3) |> pull(estimate)
+coefs
+
+coefs_sim <- rmvnorm(n = 100, mean = coefs, sigma = vcov(m3)) |> 
+  as_tibble() |> 
+  set_names(tidy(m3) |> mutate(term = paste0(term, "_beta")) |> pull(term)) |> 
+  mutate(sim_round = row_number())
+
+new_data <- perc_df |> 
+  mutate(trade_dep_total = 0) |> 
+  bind_rows(
+    perc_df |> 
+      mutate(trade_dep_total = mean(full_df$trade_dep_total) * 100)
+  ) |> 
+  group_by(trade_dep_total) |> 
+  mutate(id = row_number()) |> 
+  select(id, mid_onset_all, trade_dep_total, conttype, v2x_polyarchy1, v2x_polyarchy2)
+
+sim_data <- coefs_sim |> 
+  mutate(sim_round = row_number()) |> 
+  full_join(new_data, by = character()) |> 
+  mutate(
+    .fitted = `(Intercept)_beta` +
+      trade_dep_total_beta * trade_dep_total +
+      conttype_beta * conttype +
+      v2x_polyarchy1_beta * v2x_polyarchy1 +
+      v2x_polyarchy2_beta * v2x_polyarchy2,
+    .fitted = plogis(.fitted)
+  ) |> 
+  arrange(sim_round, id, trade_dep_total)
+
+result <- sim_data |>
+  group_by(sim_round, id) |> 
+  mutate(diff = (.fitted - lag(.fitted)) * 100) |> 
+  drop_na(diff) |> 
+  ungroup() |> 
+  summarise(lower_bound = quantile(diff, 0.025),
[…]

### 2022-12-13 data-processing/data-processing.R

@@ -6,8 +6,11 @@ library(jsonlite)
 
 # Scope -------------------------------------------------------------------
 
+min_year <- 2014
+max_year <- 2014
+
 scope <- create_dyadyears(system = "cow", mry = F, directed = T) |> 
-  filter(year == 2014, !ccode1 %in% c(347, 713)) |> 
+  filter(year %in% min_year:max_year, !ccode1 %in% c(347, 713)) |> 
   mutate(
     country1 = countrycode(ccode1, "cown", "country.name"),
     country2 = countrycode(ccode2, "cown", "country.name")
@@ -47,7 +50,7 @@ trade_raw <- rio::import(here::here("data-raw", "trade_data_2014.csv")) |>
 trade_bilat <- trade_raw |> 
   filter(country2 != "World")
 
-gdp_df <- wbstats::wb_data("NY.GDP.MKTP.CD", start_date = 2014, end_date = 2014, return_wide = F) |> 
+gdp_df <- wbstats::wb_data("NY.GDP.MKTP.CD", start_date = min_year, end_date = max_year, return_wide = F) |> 
   transmute(
     country1 = countrycode(iso3c, "iso3c", "country.name"), 
     year = date, 
@@ -74,14 +77,7 @@ trade_dep_max <- trade_dep |>
   group_by(country1, country2, year) |> 
   slice_max(trade_dep_total, with_ties = F) |> 
   ungroup() |> 
-  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
-
-trade_dep_75 <- trade_dep |> 
-  group_by(country1, country2, year) |> 
-  mutate(trade_dep_75 = quantile(trade_dep_total)[[4]]) |> 
-  view()
-  ungroup() |> 
-  select(year:cmd_code, cmd_desc_e, trade_dep_max = trade_dep_total)
+  select(year:cmd_code, cmd_desc_e, trade_dep_total)
 
 trade_dep_broad <- trade_bilat |> 
   group_by(year, country1, country2) |> 
